Making Sequential Programs Deterministic

### 2019-03-26 design.tex
maps and sets discussion

\subsection{Maps discussion}\label{maps}

\begin{itemize}
    \item A map is deterministic iff the \<entrySet()> is
          deterministic; equivalently, iff the \<keySet()> and \<values()>
          are both deterministic
    \item the iteration order of a keyset is the same as the iteration order of entrySet and Values.
    \item \<HashMap> - can be \<@OrderNonDet> or \<@NonDet>.
    \item \<LinkedHashMap> - can be \<@Det>, \<@OrderNonDet>, ot \<@NonDet>.
    \item \<TreeMap> - can be \<@Det> or \<@NonDet>.
    \item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
    \item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
    \item It was probably a bad design choice to make HashMap a subclass of Map (talk about how
    every method is overriden). Similarly for HashSet.
\end{itemize}

\begin{itemize}
    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
    We explicitly annotate the return type on constructors as
    \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
    HashMap>.
\end{itemize}

>>>>

\subsection{Maps discussion}\label{maps}
A \<Map> is a set of mappings from keys to values. A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.

We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
\begin{itemize}
    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered according
    to the natural ordering of the keys. 
\end{itemize}

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
Figure~\ref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of Figure~\ref{fig-creation-rules}.

\begin{figure}
    $\infer[\sc{\mathit{HASHMAP \ CREATION}}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$

    $\infer[\sc{\mathit{TREEMAP \ CREATION}}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$

    $\infer[\sc{\mathit{HASHSET \ CREATION}}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$

    $\infer[\sc{\mathit{TREESET \ CREATION}}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
    \caption{\<Map> and \<Set> creation rules}
    \label{fig-creation-rules}
\end{figure}

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Blochs book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

\begin{itemize}
    \item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
    \item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
\end{itemize}
